### 2020-04-05 06b-bisimulation.tex
notazione formule
nella sezione 6.3 si confondeva l'entità G con la formula G. Ho usato \mathsf{} per tutte le formule.

\begin{definition}[Assertion Language]
\label{def:assetionl}
Assertions are built from: 
$$
%B ::=  s~\mid~ \tilde{s}~\mid~ \hat{s} ~\mid~ \overline{s} ~\mid~  \underline{s} ~\mid~ F \oplus F  
\begin{array}{lcl}
\zeta & ::= & 
\alpha ~\mid ~ 
? ~\mid ~ 
[N]
\\
F & ::= & 
\epsilon ~\mid ~ 
\zeta ~\mid ~ 
F::F ~\mid ~ 
F^+ ~\mid ~ 
F^* ~\mid ~ 
F \vee F ~\mid ~ 
F \wedge F
\end{array}
$$
%\noindent where $s \in {\cal C}$, and  $\oplus \in \{\wedge, \vee, \to \}$. 
where $\alpha \in \mathit{syms}$ and $N\subseteq \mathit{syms}$.
%, and  $\oplus \in \{\wedge, \vee \}$. 
% \{\wedge, \vee, \to \}$. 
\end{definition}

@@

An \emph{assertion} $F$ is either the empty string $\epsilon$, a singleton assertion $\zeta$, the concatenation of two assertions $F_1::F_2$, the replication of $F$ for 1 or more times $F^+$, the replication of $F$ for 0 or more times $F^*$, the disjunction of two assertions $F_1 \vee F_2$ or their conjunction $F_1 \wedge F_2$. We write $\star$ as an abbreviation of $?^*$.

@@

\begin{definition}[Semantics of Assertions]
 We define $\llbracket F\rrbracket \subseteq\wp(\mathit{syms}^*)$ by induction on the structure of $F$: 
\[
\begin{array}{rcl@{\hspace{1cm}}rcl}
 \llbracket \epsilon \rrbracket &\defeq&  \{\epsilon\}
 & \llbracket F_1::F_2 \rrbracket &\defeq& \{\omega_1 :: \omega_2~\mid~ \omega_1 \in \llbracket F_1 \rrbracket \wedge \omega_2 \in \llbracket F_2 \rrbracket\}  \\
 \llbracket \alpha \rrbracket &\defeq&  \{\alpha\} 
 & \llbracket F^+ \rrbracket &\defeq& \llbracket F \rrbracket^+  \\
 \llbracket ? \rrbracket &\defeq&   \mathit{syms}
 & \llbracket F^* \rrbracket &\defeq& \llbracket F \rrbracket^*  \\
 \llbracket [N]\rrbracket &\defeq&   N
 & \llbracket  F_1 \vee F_2 \rrbracket &\defeq& \llbracket  F_1\rrbracket \cup \llbracket  F_2\rrbracket  \\
 &&&\llbracket  F_1 \wedge F_2 \rrbracket &\defeq& \llbracket F_1 \rrbracket \cap \llbracket F_2\rrbracket 
\end{array}
\]
\end{definition}

\begin{definition}[Satisfaction as Membership]\label{sec:semantics}
Let  $\upsilon$ be  a transition label, and $F$ be an assertion.
We write $\upsilon \entails F$ (read as the transition label $v$ satisfies the assertion $F$) 
if $\mathit{flat}(\upsilon)\in \llbracket F\rrbracket$, otherwise we write $\upsilon \not\entails F$ (or also $\upsilon \entails \neg F$) and say that $F$ does not hold at $v$. 
\end{definition}

Given two transition labels $v,w$ we write $v\equiv_F w$ if $v \entails F\ \Leftrightarrow\ w \entails F$, i.e. if both $v,w$ satisfy $F$ or they do not.

@@

\begin{definition}[Bio-similarity  $ \sim_F $]
Given an assertion $F$, a \emph{bio-simulation} $\mathbf{R}_F$ that respects $F$ is a binary relation over c\CNA \  processes such that, whenever $P~\mathrel{\mathbf{R}_F}~Q$ then:
 \begin{itemize} 
 \item
 for any $v,P'$ such that 
 $P \xRightarrow{v} P'$ then there exist $w,Q'$ such that $Q  \xRightarrow{w} Q'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\item
 for any $w,Q'$ such that 
 $Q \xRightarrow{w} Q'$ then there exist $v,P'$ such that $P  \xRightarrow{v} P'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\end{itemize}
%\item i$Q \xrightarrow{v} Q'$ with $v \entails F$ iff $\exists v',P'$ s.t.~$v \stretcheq v'$, $P  \xrightarrow{v'} P'$ with $v \entails F$. and $P' \mathrel{\mathbf{R}_{F}} Q'$.
%\end{itemize}
We let $ \sim_F $ denote the largest  bio-simulation and we say that $P$ is \emph{bio-similar} to $Q$, with respect to $F$, if $P \sim_F Q$.
\end{definition}

@@

\begin{remark}
An alternative way to look at a bio-simulation that respects $F$ is to define it as an ordinary bisimulation over the transition system labelled over $\{F,\neg F\}$ obtained by transforming each transition $P \xRightarrow{v} P'$ such that $v\entails F$ into $P \xrightarrow{F} P'$ and  each transition $P \xRightarrow{v} P'$ such that $v\not\entails F$ into $P \xrightarrow{\neg F} P'$.
\end{remark}

@@

Now, we introduce a slightly modified version of the Hennessy Milner Logic~\cite{HM80}, called bioHML; due to the reasons we explained above, we do not want to look at the complete transition labels, thus we rely on our simple assertion language to make it parametric to the assertion $F$ of interest:
\begin{definition}[BioHML]
Let   $F$ be an assertion, then 
the set of bioHML formulas $G$ that respects $F$ are built by the following syntax:
$$
\begin{array}{rcl}
\chi & ::= & F~\mid~ \neg F\\
G &::= &{\tt t} ~\mid~ {\tt f} ~\mid~ G\wedge G ~\mid~ G \vee G ~\mid~ \langle \chi \rangle G ~\mid~ [\chi]G
\end{array}
$$
\end{definition}

\begin{remark}
An alternative way to look at bioHML formulas is as ordinary HML formulas over the set of labels $\{F,\neg F\}$.
\end{remark}

@@

\begin{definition}[Semantics of BioHML]
 We define $\llbracket G\rrbracket \subseteq {\cal P}$ by induction on the structure of $G$:
 %(read as the transition label $v$ satisfies the formula $F$) as: 
\[
\begin{array}{ccl@{\hspace{1cm}}ccl}
 \llbracket {\tt t}\rrbracket &\defeq& {\cal P} &  \llbracket G\wedge H\rrbracket &\defeq&  \llbracket G\rrbracket  \cap \llbracket H\rrbracket  \\
\llbracket {\tt f}\rrbracket& \defeq& \emptyset &
  \llbracket G\vee H\rrbracket &\defeq&  \llbracket G\rrbracket  \cup \llbracket H\rrbracket \\
%\item   $ v \entails F_1 \vee F_2$ iff   $v \entails F_1$  $\vee$  $v \entails F_2$.
\end{array}
\]
\[
\begin{array}{lcl}
 \llbracket  \langle \chi \rangle G \rrbracket &\defeq& \{P \in {\cal P}: \exists v,P'.\ P\xRightarrow{v}P' \mbox{ with } v \entails \chi \mbox{ and } P' \in  \llbracket  G \rrbracket\} \\
 \llbracket [\chi]G \rrbracket &\defeq& \{P \in {\cal P}: \forall v,P'.\ P\xRightarrow{v}P' \mbox{ implies } v \entails \chi \mbox{ and } P' \in  \llbracket  G \rrbracket\}
\end{array}
\]
We write $P \entails G$ ($P$ satisfies $G$) if and only if $P \in  \llbracket G \rrbracket$.
\end{definition}

Negation is not included in the syntax, but the converse $\overline{G}$ of a bioHML formula $G$ can be easily defined inductively in the same way as for HML logic.

\begin{definition}[Converse]
Given a bioHML formula $G$ we define its converse $\overline{G}$ as follows:
\[
\begin{array}{rcl@{\hspace{1cm}}rcl@{\hspace{1cm}}rcl}
\overline{\tt t} & \defeq & {\tt f} &
\overline{G \wedge H} & \defeq & \overline{G} \vee \overline{H} &
\overline{\langle \chi \rangle G} & \defeq & [\chi]\overline{G}
\\
\overline{\tt f} & \defeq & {\tt t} &
\overline{G \vee H} & \defeq & \overline{G} \wedge \overline{H} &
\overline{[\chi]G} & \defeq & \langle \chi \rangle\overline{G}
\end{array}
\]
\end{definition}

We observe that, as expected, for any bioHML formula $G$ and process $P$ we have $\overline{\overline{G}} = G$ and  $P \entails \overline{G}$ iff  $P\not\entails G$.

We let $\mathcal{L}_F$ be the set of all bioHML formulas that respects $F$ and 
we say that two processes $P,Q$ are bio-logically equivalent w.r.t. $F$, written $P \equiv_{\mathcal{L}_F}Q$, when $P$ and $Q$ satisfy the exactly the same bioHML formulas in $\mathcal{L}_F$, i.e. when for any $G\in \mathcal{L}_F$ we have $P \entails G\ \Leftrightarrow Q \entails G$.

@@

$\sim_F\ =\ \equiv_{\mathcal{L}_F}$

>>>>

\begin{definition}[Assertion Language]
\label{def:assetionl}
Assertions are built from: 
$$
%B ::=  s~\mid~ \tilde{s}~\mid~ \hat{s} ~\mid~ \overline{s} ~\mid~  \underline{s} ~\mid~ F \oplus F  
\begin{array}{lcl}
\zeta & ::= & 
\alpha ~\mid ~ 
? ~\mid ~ 
[N]
\\
\mathsf{F} & ::= & 
\epsilon ~\mid ~ 
\zeta ~\mid ~ 
\mathsf{F}::\mathsf{F} ~\mid ~ 
\mathsf{F}^+ ~\mid ~ 
\mathsf{F}^* ~\mid ~ 
\mathsf{F} \vee \mathsf{F} ~\mid ~ 
\mathsf{F} \wedge \mathsf{F}
\end{array}
$$
%\noindent where $s \in {\cal C}$, and  $\oplus \in \{\wedge, \vee, \to \}$. 
where $\alpha \in \mathit{syms}$ and $N\subseteq \mathit{syms}$.
%, and  $\oplus \in \{\wedge, \vee \}$. 
% \{\wedge, \vee, \to \}$. 
\end{definition}

@@

An \emph{assertion} $\mathsf{F}$ is either the empty string $\epsilon$, a singleton assertion $\zeta$, the concatenation of two assertions $\mathsf{F}_1::\mathsf{F}_2$, the replication of $\mathsf{F}$ for 1 or more times $\mathsf{F}^+$, the replication of $\mathsf{F}$ for 0 or more times $\mathsf{F}^*$, the disjunction of two assertions $\mathsf{F}_1 \vee \mathsf{F}_2$ or their conjunction $\mathsf{F}_1 \wedge \mathsf{F}_2$. We write $\star$ as an abbreviation of $?^*$.

@@

\begin{definition}[Semantics of Assertions]
 We define $\llbracket \mathsf{F}\rrbracket \subseteq\wp(\mathit{syms}^*)$ by induction on the structure of $\mathsf{F}$: 
\[
\begin{array}{rcl@{\hspace{1cm}}rcl}
 \llbracket \epsilon \rrbracket &\defeq&  \{\epsilon\}
 & \llbracket \mathsf{F}_1::\mathsf{F}_2 \rrbracket &\defeq& \{\omega_1 :: \omega_2~\mid~ \omega_1 \in \llbracket \mathsf{F}_1 \rrbracket \wedge \omega_2 \in \llbracket \mathsf{F}_2 \rrbracket\}  \\
 \llbracket \alpha \rrbracket &\defeq&  \{\alpha\} 
 & \llbracket \mathsf{F}^+ \rrbracket &\defeq& \llbracket \mathsf{F} \rrbracket^+  \\
 \llbracket ? \rrbracket &\defeq&   \mathit{syms}
 & \llbracket \mathsf{F}^* \rrbracket &\defeq& \llbracket \mathsf{F} \rrbracket^*  \\
 \llbracket [N]\rrbracket &\defeq&   N
 & \llbracket  \mathsf{F}_1 \vee \mathsf{F}_2 \rrbracket &\defeq& \llbracket  \mathsf{F}_1\rrbracket \cup \llbracket  \mathsf{F}_2\rrbracket  \\
 &&&\llbracket  \mathsf{F}_1 \wedge \mathsf{F}_2 \rrbracket &\defeq& \llbracket \mathsf{F}_1 \rrbracket \cap \llbracket \mathsf{F}_2\rrbracket 
\end{array}
\]
\end{definition}

\begin{definition}[Satisfaction as Membership]\label{sec:semantics}
Let  $\upsilon$ be  a transition label, and $\mathsf{F}$ be an assertion.
We write $\upsilon \entails \mathsf{F}$ (read as the transition label $v$ satisfies the assertion $\mathsf{F}$) 
if $\mathit{flat}(\upsilon)\in \llbracket \mathsf{F}\rrbracket$, otherwise we write $\upsilon \not\entails \mathsf{F}$ (or also $\upsilon \entails \neg \mathsf{F}$) and say that $\mathsf{F}$ does not hold at $v$. 
\end{definition}

Given two transition labels $v,w$ we write $v\equiv_{\mathsf{F}} w$ if $v \entails \mathsf{F}\ \Leftrightarrow\ w \entails \mathsf{F}$, i.e. if both $v,w$ satisfy $\mathsf{F}$ or they do not.

@@

\begin{definition}[Bio-similarity  $ \sim_{\mathsf{F}}$]
Given an assertion $\mathsf{F}$, a \emph{bio-simulation} $\mathbf{R}_{\mathsf{F}}$ that respects $\mathsf{F}$ is a binary relation over c\CNA \  processes such that, whenever $P~\mathrel{\mathbf{R}_{\mathsf{F}}}~Q$ then:
 \begin{itemize} 
 \item
 for any $v,P'$ such that 
 $P \xRightarrow{v} P'$ then there exist $w,Q'$ such that $Q  \xRightarrow{w} Q'$ with $v\equiv_{\mathsf{F}} w$ and $P' \mathrel{\mathbf{R}_{\mathsf{F}}} Q'$.
\item
 for any $w,Q'$ such that 
 $Q \xRightarrow{w} Q'$ then there exist $v,P'$ such that $P  \xRightarrow{v} P'$ with $v\equiv_{\mathsf{F}} w$ and $P' \mathrel{\mathbf{R}_{\mathsf{F}}} Q'$.
\end{itemize}
%\item i$Q \xrightarrow{v} Q'$ with $v \entails F$ iff $\exists v',P'$ s.t.~$v \stretcheq v'$, $P  \xrightarrow{v'} P'$ with $v \entails F$. and $P' \mathrel{\mathbf{R}_{F}} Q'$.
%\end{itemize}
We let $ \sim_{\mathsf{F}} $ denote the largest  bio-simulation and we say that $P$ is \emph{bio-similar} to $Q$, with respect to $\mathsf{F}$, if $P \sim_{\mathsf{F}} Q$.
\end{definition}

@@

\begin{remark}
An alternative way to look at a bio-simulation that respects $\mathsf{F}$ is to define it as an ordinary bisimulation over the transition system labelled over $\{\mathsf{F},\neg \mathsf{F}\}$ obtained by transforming each transition $P \xRightarrow{v} P'$ such that $v\entails \mathsf{F}$ into $P \xrightarrow{\mathsf{F}} P'$ and  each transition $P \xRightarrow{v} P'$ such that $v\not\entails \mathsf{F}$ into $P \xrightarrow{\neg \mathsf{F}} P'$.
\end{remark}

@@

Now, we introduce a slightly modified version of the Hennessy Milner Logic~\cite{HM80}, called bioHML; due to the reasons we explained above, we do not want to look at the complete transition labels, thus we rely on our simple assertion language to make it parametric to the assertion $\mathsf{F}$ of interest:

\begin{definition}[BioHML]
Let   $\mathsf{F}$ be an assertion, then 
the set of bioHML formulas $\mathsf{G}$ that respects $\mathsf{F}$ are built by the following syntax:
$$
\begin{array}{rcl}
\chi & ::= & \mathsf{F}~\mid~ \neg \mathsf{F}\\
\mathsf{G},\mathsf{H} &::= &{\tt t} ~\mid~ {\tt f} ~\mid~ \mathsf{G}\wedge \mathsf{G} ~\mid~ \mathsf{G} \vee \mathsf{G} ~\mid~ \langle \chi \rangle \mathsf{G} ~\mid~ [\chi]\mathsf{G}
\end{array}
$$
\end{definition}

\begin{remark}
An alternative way to look at bioHML formulas is as ordinary HML formulas over the set of labels $\{\mathsf{F},\neg \mathsf{F}\}$.
\end{remark}

@@

\begin{definition}[Semantics of BioHML]
 We define $\llbracket \mathsf{G}\rrbracket \subseteq {\cal P}$ by induction on the structure of $\mathsf{G}$:
 %(read as the transition label $v$ satisfies the formula $F$) as: 
\[
\begin{array}{ccl@{\hspace{1cm}}ccl}
 \llbracket {\tt t}\rrbracket &\defeq& {\cal P} &  \llbracket \mathsf{G}\wedge \mathsf{H}\rrbracket &\defeq&  \llbracket \mathsf{G}\rrbracket  \cap \llbracket \mathsf{H}\rrbracket  \\
\llbracket {\tt f}\rrbracket& \defeq& \emptyset &
  \llbracket \mathsf{G}\vee \mathsf{H}\rrbracket &\defeq&  \llbracket \mathsf{G}\rrbracket  \cup \llbracket \mathsf{H}\rrbracket \\
%\item   $ v \entails F_1 \vee F_2$ iff   $v \entails F_1$  $\vee$  $v \entails F_2$.
\end{array}
\]
\[
\begin{array}{lcl}
 \llbracket  \langle \chi \rangle \mathsf{G} \rrbracket &\defeq& \{P \in {\cal P}: \exists v,P'.\ P\xRightarrow{v}P' \mbox{ with } v \entails \chi \mbox{ and } P' \in  \llbracket  \mathsf{G} \rrbracket\} \\
 \llbracket [\chi]\mathsf{G} \rrbracket &\defeq& \{P \in {\cal P}: \forall v,P'.\ P\xRightarrow{v}P' \mbox{ implies } v \entails \chi \mbox{ and } P' \in  \llbracket  \mathsf{G} \rrbracket\}
\end{array}
\]
We write $P \entails \mathsf{G}$ ($P$ satisfies $\mathsf{G}$) if and only if $P \in  \llbracket \mathsf{G} \rrbracket$.
\end{definition}

Negation is not included in the syntax, but the converse $\overline{\mathsf{G}}$ of a bioHML formula $\mathsf{G}$ can be easily defined inductively in the same way as for HML logic.

\begin{definition}[Converse]
Given a bioHML formula $\mathsf{G}$ we define its converse $\overline{\mathsf{G}}$ as follows:
\[
\begin{array}{rcl@{\hspace{1cm}}rcl@{\hspace{1cm}}rcl}
\overline{\tt t} & \defeq & {\tt f} &
\overline{\mathsf{G} \wedge \mathsf{H}} & \defeq & \overline{\mathsf{G}} \vee \overline{\mathsf{H}} &
\overline{\langle \chi \rangle \mathsf{G}} & \defeq & [\chi]\overline{\mathsf{G}}
\\
\overline{\tt f} & \defeq & {\tt t} &
\overline{\mathsf{G} \vee \mathsf{H}} & \defeq & \overline{\mathsf{G}} \wedge \overline{\mathsf{H}} &
\overline{[\chi]\mathsf{G}} & \defeq & \langle \chi \rangle\overline{\mathsf{G}}
\end{array}
\]
\end{definition}

We observe that, as expected, for any bioHML formula $G$ and process $P$ we have $\overline{\overline{\mathsf{G}}} = \mathsf{G}$ and  $P \entails \overline{\mathsf{G}}$ iff  $P\not\entails \mathsf{G}$.

We let $\mathcal{L}_{\mathsf{F}}$ be the set of all bioHML formulas that respects $\mathsf{F}$ and 
we say that two processes $P,Q$ are bio-logically equivalent w.r.t. $\mathsf{F}$, written $P \equiv_{\mathcal{L}_{\mathsf{F}}}Q$, when $P$ and $Q$ satisfy the exactly the same bioHML formulas in $\mathcal{L}_{\mathsf{F}}$, i.e. when for any $\mathsf{G}\in \mathcal{L}_{\mathsf{F}}$ we have $P \entails \mathsf{G}\ \Leftrightarrow Q \entails \mathsf{G}$.

@@

$\sim_{\mathsf{F}}\ =\ \equiv_{\mathcal{L}_{\mathsf{F}}}$
